---
title: "R: the very basics"
output: 
  revealjs::revealjs_presentation:
    css: style1.css
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 4.5, fig.align = "center", comment = "##", collapse = T)
library(tidyverse)
theme_set(theme_bw())
```

# Introduction
<<<<<<< HEAD
=======

## Content
>>>>>>> ec5eaddab2ce6e7fa3c29ff9ca3b8ee807a432f7

## <a href="https://www.r-project.org/">R</a>

Code-based statistical software

Developed in the 1990s from *S*

Free of cost and free to develop further

Add-ons through user-developed *packages*, published on *CRAN* (the Comprehensive R Archive Network)

### <a href="https://rstudio.com/">RStudio</a>

The most popular interface to R

# Objects and functions

## Objects

Any collection of data used in R can be assigned a name and stored as an *object* using the assign arrow (`<-` or `->`)

<<<<<<< HEAD
=======
# Objects and functions

## Objects

Any collection of data used in R can be assigned a name and stored as an *object* using the assign arrow (`<-` or `->`)

>>>>>>> ec5eaddab2ce6e7fa3c29ff9ca3b8ee807a432f7
```{r}
a <- c(1,2,3) 
b <- data.frame(Name = c("Abby", "Bobby"), 
                Height = c(160,158))
```
<<<<<<< HEAD
=======

## Data mode

In R, single data 
>>>>>>> ec5eaddab2ce6e7fa3c29ff9ca3b8ee807a432f7

## Data structure

Objects have a *data structure*

- `a` is a vector (single dimension)

```{r}
a
```

- `b` is a data frame (two dimensions)

```{r}
b
```

<<<<<<< HEAD
A data frame can be seen as a collection of vectors

## Class of a vector

The *class* of a vector is decided by the type of data it contains

- `c(1,2,3)` is a *numeric vector*
- `c("Aluminium", "Barium", "Curium")` is a *character vector*

The class impacts the behaviour of vectors

```{r, eval = F}
c(1,2,3) + 4 # Returns 5,6,7

c("1", "2", "3") + 4 # Returns an error: non-numeric argument to binary operator
```

*Factors* are vectors which have a determined number of possible levels

## Subsetting

=======
## Subsetting

>>>>>>> ec5eaddab2ce6e7fa3c29ff9ca3b8ee807a432f7
Possible to extracts subsets of data using square brackets `[]`

```{r}
a[2:3] # Extract elements 2 and 3 from vector a
b[1, 2] # Extract row 1 and column 2 of data frame b

b$Name # Extract the Name column of b
```

## Functions

R operates by applying functions to objects

The `mean` function calculates the mean value, so `mean(a)` gives the mean of the vector `a`

```{r}
a # Prints the object a
mean(a) # Prints the mean of a
```

# A first example

## Example data

R and add-on packages come with a large library of example data

The `Animals` dataset from the `MASS` package contains body mass (kg) and brain mass (g) of 28 animals

```{r}
data(Animals, package = "MASS")
dat <- Animals # Renames to dat

dat[1:5, 1:2] # Prints rows 1 to 5 and columns 1 to 2
```

## Transforming the data

Set both measures to the same scale

```{r}
dat$body <- dat$body * 1000 # Transform body from kg to g
```

Calculate a brain-to-body ratio and store as new column

```{r}
dat$brain_to_body <- dat$brain / dat$body # Divides brain mass by body mass
```

## Plot data

```{r}
plot(dat, main = "Scatterplots of mass") # Plotting using the basic plot function
```

# Data import and handling

## Data import

Exist functions for most common data formats

The choice of function depends on the file format

```{r, eval = F}
# CSV
dat1 <- read.csv("filename.csv", sep = ";", dec = ".")
# Import from csv with semicolon as separator and dot as decimal point

# Excel
library(readxl)
dat2 <- read_excel("filename.xlsx", sheet = "Sheet 2", range = "A3:C13")
# Import from Excel file with data in cells A3 to C13 in sheet two
```

This requires that the file is in the working directory of the R session
<<<<<<< HEAD

The working directory can be set using `set.wd`

## Data handling

Common operations

- Select columns
- Filter rows (subsetting)
- Transform variables
- Summarise by a category

## Base functions

Square brackets can be used for both selecting columns and filtering rows

```{r, eval = F}
dat[, c("body", "brain")] # Selects columns brain and body
dat[dat$body > 1500, ] # Filters on body mass greater than 100
```

=======

The working directory can be set using `set.wd`

## Data handling

Common operations

- Select columns
- Filter rows (subsetting)
- Transform variables
- Summarise by a category

## Base functions

Square brackets can be used for both selecting columns and filtering rows

```{r, eval = F}
dat[, c("body", "brain")] # Selects columns brain and body
dat[dat$body > 1500, ] # Filters on body mass greater than 100
```

>>>>>>> ec5eaddab2ce6e7fa3c29ff9ca3b8ee807a432f7
Data can be transformed by adding new columns

```{r}
dat$greater_than_1500 <- dat$body > 1500 # Create new variable
dat[1:5,]
```

## Summarise by categorical variable

Data can be summarised using the `aggregate` function

```{r}
aggregate(brain ~ greater_than_1500, dat, mean)
```

Mean brain mass by body mass category


# Tidyverse
## Tidyverse and piping

The *Tidyverse* is a collection of R add-on packages with a shared structure and idea

*Tidy data*: every observation is a row and every variable is a column

Include the packages `dplyr`, `ggplot2` and `readxl`

## Piping

Tidyverse function often used along with *piping*

Piping strings functions together by using the output of one function as input in the next function

```{r}
a

a %>% mean()
```

If used well, piping can give very readable code

## Data handling revisited

```{r}
library(dplyr)

dat %>% 
  mutate(greater_than_1500 = body > 1500) %>% 
  group_by(greater_than_1500) %>% 
  summarise(Mean = mean(brain), 
            Sd = sd(brain))
<<<<<<< HEAD
```

*Take the data frame <br> and create a new categorical variable <br> and group by that variable <br> and summarise brain mass by that grouping*

## Plots in `ggplot2`

`ggplot2` is a commonly used package to produce plots

- Data is specified as *aesthetics* in an `aes` function
- Visual elements are added as *geometries* using functions starting in `geom_`

## `ggplot2` example

```{r, message = F, fig.height = 4}
library(ggplot2)

g <- ggplot(dat, aes(log(brain), log(body), color = greater_than_1500)) + 
  geom_point() +
  geom_smooth(method = "lm", alpha = 0.2)
g
```

## Interactive graphs using `plotly`

```{r, eval = F}
library(plotly)
ggplotly(g)
```

<iframe src="Figurer/graph1.html" height=400 width=800></iframe>

# Functions

## Using functions

Functions take some input (*argument*) and produce an output

Functions are applied by the function name followed by the arguments in brackets

```{r, fig.height=3}
par(mfrow = c(1,2)) # Sets the plotting window to 2-by-1
plot(1:3, col = "black", main = "Black circles")
plot(1:3, col = "green", main = "Green circles")
```

## Reading functions and help pages

Writing the function name prints the code of the function

This can be a difficult read

```{r}
sd
```

The help page of a function can be found by writing `?` followed by the function name

```{r, eval = F}
?sd
```

## Writing functions

Easy to define your own functions

```{r}
add_two <- function(x){
  x + 2
}

add_two(5)

add_inputs <- function(x, y){
  x + y
}

add_inputs(5, 2)
```

Useful when the same piece of code must be run multiple times

## Storing and extracting output

The output of a function can be assigned a name and stored

```{r}
m <- mean(a)
m
```

This allows one to apply further functions

The `lm` function estimates the regression line of brain mass on body mass and the `anova` function print the Anova table

```{r}
=======
```

*Take the data frame <br> and create a new categorical variable <br> and group by that variable <br> and summarise brain mass by that grouping*

## Plots in `ggplot2`

`ggplot2` is a commonly used package to produce plots

- Data is specified as *aesthetics* in an `aes` function
- Visual elements are added as *geometries* using functions starting in `geom_`

## `ggplot2` example

```{r, message = F}
library(ggplot2)

ggplot(dat, aes(log(brain), log(body), color = greater_than_1500)) + 
  geom_point() +
  geom_smooth(method = "lm", alpha = 0.2)
```

# Functions

## Using functions

Functions take some input (*argument*) and produce an output

Functions are applied by the function name followed by the arguments in brackets

```{r, fig.height=3.5}
par(mfrow = c(1,2))
plot(1:3, col = "black")
plot(1:3, col = "green")
```

## Reading functions and help pages

Writing the function name prints the code of the function

This can be a difficult read

```{r}
sd
```

The help page of a function can be found by writing `?` followed by the function name

```{r, eval = F}
?sd
```

## Writing functions

Easy to define your own functions

```{r}
add_two <- function(x){
  x + 2
}

add_two(5)

add_inputs <- function(x, y){
  x + y
}

add_inputs(5, 2)
```

Useful when the same piece of code must be run multiple times

## Storing and extracting output

The output of a function can be assigned a name and stored

```{r}
m <- mean(a)
m
```

This allows one to apply further functions

The `lm` function estimates the regression line of brain mass on body mass and the `anova` function print the Anova table

```{r}
>>>>>>> ec5eaddab2ce6e7fa3c29ff9ca3b8ee807a432f7
mod <- lm(brain ~ body, dat)

anova(mod)
```

##

It also makes it possible to extract information about the estimated model using `$`

```{r}
mod$coefficients # Prints the estimated coefficients
mod$fitted.values # Prints the estimated values
```

# Troubleshooting

## 

### Troubleshooting

Read error messages

Google and Stack Overflow

### A common SLU problem

Problems installing packages to external SLU server

Possible to redirect to an installation folder on the local hard-drive

Guide from SLU IT available on the course page

# Scripts and output

## Scripts

RStudio makes it easy to work with scripts (i.e. collections of code):

>- Code section can be run directly from the script to the console
>- Special R project folders can be used to keep track of the working directory, scripts, raw data files, and output
>- Easy to keep scripts short by splitting into different scripts for different functionality (import, plotting, models, etc)
<<<<<<< HEAD
>- Scripts can be split into sections using hashtag, section name, and four dashes: `# Plotting functions ----`

## Markdown in R

R and RStudio include functions to write a report directly in R

Easiest way to get data from R to a Word document or a presentation

In RStudio: *New file > R Markdown... >* select a suitable file format

A great resource for your hand-ins during the course
=======
>- Scripts can be split into sections using four dashes `----`

## Markdown in R


>>>>>>> ec5eaddab2ce6e7fa3c29ff9ca3b8ee807a432f7
